WSClean: an implementation of a fast, generic wide-field imager

### 2014-02-07 articles/2014/Article-WSClean/wsclean.tex
Article almost done

New generation of wide-field observatories are producing data sets that are orders of magnitude larger than before. Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}). At the MWA, some of the data is reduced with the w-projection algorithm implementation of the Common Astronomy Software Applications (CASA, \citealt{casa}). This is an attractive option because of its many features, high quality and free availability. However, at the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with CASA's w-projection algorithm. Another option for reducing MWA data is the Real-Time System (RTS), which has been designed as an efficient calibrating and imaging pipeline specifically for MWA data \citep{rts-mwa}. It can use GPUs for improving efficiency, and uses a w-projection kernel during imaging to correct for the w-term. The kernel can be kept small, because the data is processed in snapshots that have been phase-centered to Zenith. The RTS can not perform deconvolution as of yet.

@@

The new imager, called ``WSClean'', is written in the C++ language. It reads visibilities from CASA measurement sets and writes output images to Flexible Image Transport System (FITS) files. Several steps are multi-threaded: reading and writing; gridding from and to different $w$-layers; performing the Fast Fourier transforms; and H\"ogbom Clean iterations (peak-finding and image subtraction, \citealt{hogbom-clean}). For the latter, intrinsics are used as well. Because of these optimizations, minor Cleaning iterations with an image size of $3072\times3072$ are performed at a rate of hundreds per second. This is fast enough to make the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, less relevant. However, the PSF varies for imaging with non-zero $w$-values, and subtracting a constant PSF in image space leads to inaccuracies. After a number of minor iterations, it is therefore beneficial to invert the model back to the visibilities via prediction, and subtract the model directly from the visibilities \citep{wprojection-cornwell}. This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}.

@@

For the WSClean time cost function $t_\textrm{WSClean}$ we find 
\begin{align}\label{eq:wsclean-computing-cost}
& t_\textrm{WSClean} (w_e,N_\textrm{Mvis},N_\textrm{freq},N_\textrm{kpix}) = \\ \notag%
& N_\textrm{chan} \left( 0.452 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} (w_e + 0.611) + 244 N_\textrm{Mvis} \right)
\end{align}
and for CASA we find
\begin{align}\label{eq:casa-computing-cost}
& t_\textrm{CASA}(w_e,N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = \\ \notag%
& N_\textrm{chan} \left(1.03 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} + 5.93 N_\textrm{Mvis} (w_e + 1.71)^2\right) + 172.
\label{eq:casa-computing-cost}%
\end{align}
Parameter $w_e$ is the maximum $w$ value and is estimated with
\begin{equation}
 w_e = \frac{1}{\lambda}\left[\max |\Delta \mathbf{x}| \sin \textrm{ZA} + \max \Delta z \cos \textrm{ZA} \left(1.0 - \cos(\frac{1}{2}\textrm{FOV})\right)\right],
\end{equation}
where $\max |\Delta \mathbf{x}|$ is the maximum baseline length, $\max \Delta z$ is the maximum height difference between antennas and $\lambda$ is the wavelength.

\begin{table*}
\begin{tabular}{lrrrrrrr}
 Survey   & $\lambda$ & FWHM  & Resolution & $\max |\Delta \mathbf{x}|$ & $\max \Delta z$ & int. time & B/$\delta \nu$ \\
\hline
 GLEAM    &           &       & 2'     & 2.9~km & 5~m & 2~s & 32~MHz/40~kHz\\
 EMU wide & 20 cm  & 1\degree & 10''    & 6~km &  0~m &     & ?/?\\
 MSSS low &  5 m   & 9.8\degree & 100'' & 5~km &      &10~s & 16~MHz/4~khZ\\
 MSSS high&  2 m   & 3.8\degree & 120'' & 5~km &      &10~s & 16~MHz/4~khZ\\
\end{tabular}
\end{table*}

The $w$-projection and $w$-stacking algorithms can be combined by gridding onto several $w$-layers to which large $w$-terms are corrected after the FFTs, and using a $w$-correcting kernel for making small $w$-corrections before the FFTs. This allows using small $w$-kernels and limits at the same time the number of FFTs and required memory. The AWImager has been applied like this, which let to better results compared to $w$-projection without $w$-stacking \citep{awimager-2013}. For this scenario, Eq.~\eqref{eq:casa-computing-cost} can be used to determine the optimal number of $w$-layers, or more generally the optimal distance between layers, by assuming that the cost of calculating a single $w$-layer equals the cost of imaging the data set with correspondingly smaller $w_e$ value and smaller $N_\textrm{vis}$. If $\Delta w$ is the distance between $w$-layers, then 
\begin{equation}
t(\Delta w) = \frac{w_e}{\Delta w} t_\textrm{CASA}(w'_e,N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
\end{equation}
with $N'_\textrm{vis} = \frac{N_\textrm{vis}\Delta w}{w_e}$ and $w'_e = \Delta w$.

TODO fill in formulae for common surveys, e.g. ESKAP and LOFAR.

>>>>

New generation of wide-field observatories are producing data sets that are orders of magnitude larger than before. Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}). At the MWA, some of the data is reduced with the w-projection algorithm implementation of the Common Astronomy Software Applications (CASA, \citealt{casa-2007}, \citealt{casa-2008}). This is an attractive option because of its many features, high quality and free availability. However, at the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with CASA's w-projection algorithm. Another option for reducing MWA data is the Real-Time System (RTS), which has been designed as an efficient calibrating and imaging pipeline specifically for MWA data \citep{rts-mwa}. It can use GPUs for improving efficiency, and uses a w-projection kernel during imaging to correct for the w-term. The kernel can be kept small, because the data is processed in snapshots that have been phase-centered to Zenith. The RTS can not perform deconvolution as of yet.

@@

The new imager, called ``WSClean'', is written in the C++ language. It reads visibilities from CASA measurement sets and writes output images to Flexible Image Transport System (FITS) files. Several steps are multi-threaded using the threading module of the C++11 standard library. These are: reading and writing; gridding from and to different $w$-layers; performing the Fast Fourier transforms; and H\"ogbom Clean iterations (peak-finding and image subtraction, \citealt{hogbom-clean}). For the latter, intrinsics are used as well. Because of these optimizations, minor Cleaning iterations with an image size of $3072\times3072$ are performed at a rate of hundreds per second. This is fast enough to make the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, less relevant. However, the PSF varies for imaging with non-zero $w$-values, and subtracting a constant PSF in image space leads to inaccuracies. After a number of minor iterations, it is therefore beneficial to invert the model back to the visibilities via prediction, and subtract the model directly from the visibilities \citep{wprojection-cornwell}. This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}.

@@

For the WSClean time cost function $t_\textrm{WSClean}$ we find 
\begin{align}\label{eq:wsclean-computing-cost}
& t_\textrm{WSClean} (w_e,N_\textrm{Mvis},N_\textrm{freq},N_\textrm{kpix}) = \\ \notag%
& N_\textrm{freq} \left( 0.452 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} (w_e + 0.611) + 244 N_\textrm{Mvis} \right)
\end{align}
and for CASA we find
\begin{align}\label{eq:casa-computing-cost}
& t_\textrm{CASA}(w_e,N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = \\ \notag%
& N_\textrm{freq} \left(1.03 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} + 5.93 N_\textrm{Mvis} (w_e + 1.71)^2\right) + 172.
\label{eq:casa-computing-cost}%
\end{align}
Parameter $w_e$ is the maximum $w$ value and is estimated with
\begin{equation}
 w_e = \frac{1}{\lambda}\left[\max |\Delta \mathbf{x}| \sin \textrm{ZA} + \max \Delta z \cos \textrm{ZA} \left(1.0 - \cos(\frac{1}{2}\textrm{FOV})\right)\right],
\end{equation}
where $\max |\Delta \mathbf{x}|$ is the maximum baseline length, $\max \Delta z$ is the maximum height difference between antennas and $\lambda$ is the wavelength.

\begin{table*}
\caption{Predicted imaging times for various surveys, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree.  Predictions are for $w$-projection in CASA, $w$-stacking with WSClean and $w$-stacked 12$\times$5-min snapshots with WSClean, with approximately equal accuracy.}
\begin{tabular}{lrrrrrrrrrr|rrr}
 Survey   & $\lambda$ & FWHM  & Beams & Ant & Res & $\max |\Delta \mathbf{x}|$ & $\max \Delta z$ & int. t & BW & $\Delta \nu$ & $w$-proj & $w$-stack & $w$-snaps \\
          &           &       &       &     &     & (km)                       &                 & (s)    & (MHz) &  (kHz) &  &  & \\
\hline
 GLEAM    &  2 m   & 24.7\degree& 1    & 128 & 2'   & 2.9 & 5~m & 2 & 32 & 40     &14h 02m & 1h 52 m & 1h 23m \\
 EMU      & 20 cm  & 1\degree   & 36   &  36 & 10''  & 6 &  0~m &10 & 300 & 20    &19h 49m &14h 12 m & 14h 13m\\
 MSSS low &  5 m   & 9.8\degree & 5    &  20 & 100'' & 5 &      &10 & 16 & 16     & 2h 30m & 0h 11 m & 0h 11m\\
 MSSS high&  2 m   & 3.8\degree & 5    &  40 & 120'' & 5 &      &10 & 16 & 16     & 2h 31m & 0h 38 m & 0h 39m\\
 VLSS     & 4.1 m  & 14\degree  & 1    &  27 &  80'' &11.1& 0~m &10 & 1.56 & 12.2 & 0h 32m & 0h 12 m & 0h 02m\\
\end{tabular}
\end{table*}

\subsection{Optimal snapshot time}
The computing time of snapshot imaging can be expressed in either of the $w$-correction techniques. Snapshot imaging effectively removes the dependency on ZA from $w_e$. Given the snapshot time $\Delta t_\textrm{snapshot}$ and the total time $\Delta t_\textrm{total}$, the total time cost of inversion becomes
\begin{equation}
t(\Delta t_\textrm{snapshot}) = \frac{\Delta t_\textrm{total}}{\Delta t_\textrm{snapshot}} t_\textrm{imaging}(w'_e,N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
\end{equation}
with
\begin{equation}
w'_e = \frac{1}{\lambda}\max \Delta z \left(1.0 - \cos(\frac{1}{2}(\textrm{FOV} + \angle t_\textrm{snapshot})\right)
\end{equation}
and $N'_\textrm{vis} = \frac{N_\textrm{vis}\Delta t_\textrm{total}}{t_\textrm{snapshot}}$. This function can be minimized to find the optimal snapshot time. For WSClean with MWA parameters, we find this function does not reach its minimum within 24~h, suggesting that the optimal snapshot time is as long as possible. This implies that the change in in performance due to higher $w$-values because of Earth rotation during an individual snapshot is less significant than the cost of the additional FFTs. For $w$-projection in CASA we find an optimal snapshot time of 122~seconds.

\subsection{Optimal $w$-project/$w$-stacking combination}
The $w$-projection and $w$-stacking algorithms can be combined by gridding onto several $w$-layers to which large $w$-terms are corrected after the FFTs, and using a $w$-correcting kernel for making small $w$-corrections before the FFTs. This allows using small $w$-kernels and limits at the same time the number of FFTs and required memory. The AWImager has been applied like this, which let to better results compared to $w$-projection without $w$-stacking \citep{awimager-2013}. For this scenario, Eq.~\eqref{eq:casa-computing-cost} can be used to determine the optimal number of $w$-layers, or more generally the optimal distance between layers, by assuming that the cost of calculating a single $w$-layer equals the cost of imaging the data set with correspondingly smaller $w_e$ value and smaller $N_\textrm{vis}$. If $\Delta w$ is the distance between $w$-layers, then 
\begin{equation}
t(\Delta w) = \frac{w_e}{\Delta w} t_\textrm{CASA}(w'_e,N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
\end{equation}
with $N'_\textrm{vis} = \frac{N_\textrm{vis}\Delta w}{w_e}$ and $w'_e = \Delta w.
